---
title: "Chronicler: Integration with Knitr"
author: "Lukasz A. Bartnik"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chronicler: Integration with Knitr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<script src="https://ajax.googleapis.com/ajax/libs/jquery/3.3.1/jquery.min.js"></script>

```{r setup, include=FALSE}
library(knitr)
library(utilities)

knitr::opts_chunk$set(collapse = TRUE, comment = NA, prompt = FALSE, echo = TRUE)
```

```{r hooks, echo = FALSE}
options(crayon.enabled = TRUE)
knitr::knit_hooks$set(output = utilities::ansi_handler)
knitr::knit_hooks$set(message = utilities::ansi_handler)

# try old.hooks <- set_knit_hooks(knitr::knit_hooks)

original_output_hook <- knit_hooks$get("output")
knit_hooks$set(output = utilities::create_trimming_hook(original_output_hook))
```

```{r load-chronicler,message=FALSE,include=FALSE}
options(chronicler.attach = FALSE)
library(chronicler)
```





The R snippet below adds a new engine to `knitr`. In `knitr`, an engine
is a mechanism of handling a code chunk in language other than R. In our
case, we add a very simple `chronicler` engine. From now on, each code
chunk marked as ````{chronicler}```` will be passed to the new engine.

```{r engine}
knitr::knit_engines$set(chronicler = function (options) {
  id <- as.character(options$code)
  stopifnot(is_artifact_id(id))
  
  knitted <- knit_artifact(id)
  engine_output(options, id, knitted$out, knitted$extra)
})
```

Let's see what exactly does it do. First, it casts the chunk's code as
a string (`as.character`) and immediately checks if it matches the
definition of an artifact identifier; if not, an exception is thrown.
Then it passes control to `chronicler::knit_artifact` whose output
is finally passed to `knitr::engine_output`. So what it is that
`knit_artifact` does?

It's actually pretty straightforward. It calls the default `print`
method for the `"artifact"` class and captures its output into a HTML
snippet. This is the result box which immediately follows the long
string -- our artifact identifier.

The second thing that `knit_artifact` does is capturing the sequence
of R commands that resulted in the requested artifact. It formats and
syntax-highlights each of these commands, and then puts them in another
HTML box. This additional result box that is hidden by default and can
be displayed by clicking on the `explain artifact 57f...` link. To hide
the extended explanation just click on the link once again.


```{r include=FALSE}
chronicler:::attach_to_repository(london_meters())
```

```{chronicler}
57fbe7553e11c7b0149040f5781c209b266ed637
```

