---
title: "Chronicler: Integration with ML Flow"
author: "Lukasz A. Bartnik"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chronicler: Integration with ML Flow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
library(utilities)

knitr::opts_chunk$set(collapse = TRUE, comment = NA, prompt = FALSE, echo = TRUE)
```

```{r hooks, echo = FALSE}
options(crayon.enabled = TRUE)
knitr::knit_hooks$set(output = utilities::ansi_handler)
knitr::knit_hooks$set(message = utilities::ansi_handler)

# try old.hooks <- set_knit_hooks(knitr::knit_hooks)

original_output_hook <- knit_hooks$get("output")
knit_hooks$set(output = utilities::create_trimming_hook(original_output_hook))
```

```{r load-chronicler,message=FALSE,include=FALSE}
options(chronicler.attach = FALSE)
library(chronicler)
```

# ML Flow

`ML Flow` introduces the notions of __runs__ and  __experiments__. A __run__
is a single execution of an arbitrary program that, using `ML Flow`'s API,
registers a model along with its parametrization and a number of artifacts:
plots, printouts, data sets. An __experiment__ is a group of __runs__.

`chronicler` does not have an explicit notion of __run__ or __experiment__
but we can always find a suitable mapping. Since a __run__ contains
a model, let's map each model collected by `chronicler` to a separate
__run__. Furthermore, we will examine the sequence of R commands leading
up to each of these models and extract or scalar values used in those
commands: these will be the parameters reported to `ML Flow`. Finally,
all artifacts downstream from each model will become `ML Flow` artifacts.

# Example

Once again, let's assume we run the `london_meters()` example.

```{r include=FALSE}
chronicler:::attach_to_repository(london_meters())
```


The `find_experiments()` function searches for all artifacts that match
the definition of a model and presents them together with their
parametrization and downstream products. In our example there is one
such model.

```{r}
find_experiments()
```

Now all we need to do is build a function that maps the output of
`find_experiments()` to `ML Flow` objects.

```{r}
register_with_mlflow <- function (experiment) {
  with(mlflow_start_run(), {
    # extract parameters...
    prms <- unlist(lapply(experiment$path, `[[`, i = 'parameters'))
    cnt  <- counter(1)

    # ...log them and make sure they are all named
    imap(prms, function (value, name) {
      if (!nchar(name)) name <- paste0('parameter_', counter())
      mlflow_log_param(name, value)
    })

    # log the model
    mlflow_save_model(experiment$model)
    
    # finally, log all the downstream artifacts
    mlflow_log_artifact() # TODO finish this
  })
}
```

All that is now left to do is to call `register_with_mlflow` for each
experiment in the repository.

```{r eval=FALSE}
lapply(find_experiments(), register_with_mlflow)
```
